Density-dependent selection and the limits of relative fitness

### 2018-08-31 manuscript.tex
l

Relative fitness is often parameterized in terms of selection coefficients which represent the advantages of different types relative to each other. For instance, in continuous time $s=r_2-r_1$ is the selection coefficient of type $2$ relative to type $1$. Assuming that only $2$ and $1$ are present, the change in frequency can be written as
\begin{equation}
\frac{d p_2}{dt}=s p_2 (1-p_2). \label{eq:canonicalcont}
\end{equation}
Thus, if $r_1$ and $r_2$ are constant, the frequency of the second type will grow logistically with a constant rate parameter $s$. We then say that selection is independent of frequency and density. The discrete time case is more complicated. Defining the selection coefficient by $W_2=(1+s)W_1$, and again assuming $1$ and $2$ are the only types present, we have  
\begin{equation}
\Delta p_2=\frac{W_2-W_1}{\overline{W}} p_2 (1-p_2)=\frac{s}{1+sp_2} p_2 (1-p_2). \label{eq:canonicaldisc}
\end{equation}
Hence, even in the simplest case that $W_1$ and $W_2$ are constant, selection is frequency-dependent in discrete time (note that this frequency dependence is negligible when $s$ is small compared to $1$; see \citealt{frank_2011}). We will refer to both the continuous and discrete time selection equations \eqref{eq:canonicalcont} and \eqref{eq:canonicaldisc} throughout this paper, but the simpler continuous time case will be our point of comparison for the rest of this section.

In a constant environment, and in the absence of crowding, $r_i$ is a constant ``intrinsic'' population growth rate. The interpretation of Eq.~\eqref{eq:canonicalcont} is then simple: the selection coefficient $s$ is simply the difference in intrinsic growth rates. However, growth cannot continue at a non-zero constant rate indefinitely: the population is not viable if $r_i<0$, whereas  $r_i>0$ implies endlessly increasing population density. Thus, setting aside unviable populations, the increase in population density must be checked by crowding. This implies that the Malthusian parameters $r_i$ eventually decline to zero (e.g. \citealt[pp. 203]{begon_1990}). Selection can then be density-dependent, and indeed this is probably not uncommon, because crowded and uncrowded conditions can favor very different traits \citep{travis_2013}. Eq.~\eqref{eq:canonicalcont} is then not a complete description of selection --- it lacks an additional coupled equation describing the dynamics of $N$, on which $s$ in Eq.~\eqref{eq:canonicalcont} now depends. In general we cannot simply specify the dynamics of $N$ independently, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{travis_2013}. Thus, in the presence of density-dependent selection, the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates. Note that frequency-dependent selection does not raise a similar problem, because a complete description of selection still only requires us to model the type frequencies, not the ecological variable $N$ as well.

In practice, many population genetic models simply ignore density dependence and assign a constant relative fitness to each type. Selection is typically interpreted as operating through viability, but the ecological processes underlying the regulation of population density are frequently left unspecified (e.g. \citealt{gillespie_2004}; \citealt{nagylaki_1992}; \citealt{ewens_2004}). Density either does not enter the model at all, or if finite-population size effects (``random genetic drift'') are important, then $N$ is assumed to have reached some fixed equilibrium value (Fig.~\ref{fig:Ksel}b).

@@

The constant-$N$, constant-$s$ description of selection also precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction. A variety of approaches have been developed to address this in quantitative genetics \citep{burger1995evolution,engen_2013}, population genetics \citep{bertram2017predicting} and adaptive dynamics \citep{ferriere2013eco,dieckmann2004adaptive}. Although density-dependent selection is pertinent to this longer-term issue, our focus here is the description of the time-dependent process by which selection changes allele frequencies. This is particularly critical for making sense of evolution at the genetic level, for which we now have abundant data.

@@

Here we argue that the simplified models of density-regulated growth mentioned above are misleading in their representation of the interplay between selection and density. This ultimately derives from their failure to account for ``reproductive excess'', that is, an excess of juveniles that experience stronger selection than their adult counterparts \citep{turner1968population}. By allowing selection to be concentrated at a juvenile ``bottleneck'', reproductive excess makes it possible for the density of adults to remain constant even under strong selection. Reproductive excess featured prominently in early debates about the regulation of population density (e.g. \citealt{nicholson_1954}), and also has a long history in evolutionary theory, particularly related to Haldane's ``cost of selection'' \citep{haldane_1957,turner1968population}. Additionally, reproductive excess is implicit in foundational evolutionary-genetic models like the Wright-Fisher, where each generation involves the production of an infinite number of zygotes, of which a constant number $N$ are sampled to form the next generation of adults. Likewise in the Moran model, a juvenile is always available to replace a dead adult every iteration no matter how rapidly adults are dying, and as a result $N$ remains constant.

@@

In models with detailed age structure, it is usually assumed that the density of a ``critical age group'' mediates the population's response to crowding \citep[pp. 54]{charlesworth_1994}. Reproductive excess is a special case corresponding to a critical pre-reproductive age group. A central result of the theory of density-regulated age-structured populations is that selection proceeds in the direction of increasing equilibrium density in the critical age group \citep[pp. 148]{charlesworth_1994}. This is a form of the classical $K$-selection ideas discussed above, but restricted to the critical age group (juveniles, in this case). The interdependence of pre-reproductive selection and reproductive density is thus overlooked as a result of focusing on density in the critical age group.

@@

We restrict our attention to asexual haploids, since it is then clearer how the properties of selection are tied to the underlying population ecological assumptions. We assume that reproductively mature individuals (``adults'') require their own territory to survive and reproduce. All territories are identical, and the total number of territories is $T$. Time advances in discrete iterations, each representing the time from birth to reproductive maturity. In a given iteration, the number of adults of the $i$'th type will be denoted by $n_i$, the total number of adults by $N=\sum_i n_i$, and the number of unoccupied territories by $U=T-N$. We assume that the $n_i$ are large enough that stochastic fluctuations in the $n_i$ (drift) can be ignored (with $T$ also assumed large to allow for low type densities $n_i/T\ll 1$).

@@

We assume that adults cannot be ousted by juveniles, so that recruitment to adulthood occurs exclusively in unoccupied territories. When multiple propagules land on the same unoccupied territory, the winner is determined by lottery competition: type $i$ wins a territory with probability $c_i x_i/\sum_i c_i x_i$, where $c_i$ is a constant representing relative competitive ability (Fig. \ref{fig:lottery}). Since the expected fraction of unoccupied territories with propagule composition $x_1,\ldots,x_G$ is $p_1(x_1)\cdots p_G(x_G)$ where $G$ is the number of types present, and type $i$ is expected to win a proportion $c_i x_i/\sum_i c_i x_i$ of these, type $i$'s expected territorial acquisition is given by
\begin{equation}
\Delta_+ n_i=U\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_i c_i x_i} p_1(x_1)\cdots p_G(x_G). \label{eq:growthsumuncoupled}
\end{equation}
Here the sum only includes territories with at least one propagule present. Since we do not consider random genetic drift here, we will not analyze the fluctuations around these two expectations.

@@

Here we evaluate the expectation in Eq.~\eqref{eq:growthsumuncoupled} to better understand the dynamics of density-dependent lottery competition. Similarly to the classic lottery model, we replace the $x_i$, which take different values in different territories, with ``effective'' mean values. However, since we want to allow for low propagule densities, we cannot simply replace the $x_i$ with the means $l_i$ as in the classic lottery. For a low density type, growth comes almost entirely from territories with $x_i=1$, for which its mean density $l_i\ll 1$ is not representative. We therefore separate Eq.~\eqref{eq:growthsumuncoupled} into $x_i=1$ and $x_i>1$ components, taking care to ensure that the effective mean approximations for these components are consistent with each other (details in Appendix B). The resulting variable-density approximation only requires that there are no large discrepancies in competitive ability (i.e. we do not have $c_i/c_j\gg 1$ for any two types). We obtain
\begin{equation}
\Delta_+ n_i\approx \left[e^{-L}+(R_i+A_i)\frac{c_i}{\overline{c}}\right]l_i U, \label{eq:master}
\end{equation}
where
\begin{equation}
R_i=\frac{\overline{c}e^{-l_i}(1-e^{-(L-l_i)})}{c_i +\frac{\overline{c}L- c_il_i}{L-l_i}\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}},\nonumber \label{eq:Dr}
\end{equation}
and
\begin{equation}
A_i=\frac{\overline{c}(1-e^{-l_i})}{\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}c_il_i+\frac{\overline{c}L- c_il_i}{L-l_i}\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right)}. \nonumber \label{eq:Da}
\end{equation}

@@

We now turn to the issue of how density changes as a consequence of selection in our variable-density lottery, and in previous models of selection in density-regulated populations. In the latter, selection under crowded conditions typically induces changes in equilibrium density (see Introduction). In our variable-density lottery model, however, the competitive ability trait $c$ is not density-regulating, even though $c$ contributes to fitness under crowded conditions. Consequently, $c$-selection does not cause density to change. In this section we compare this $c$-selection behavior with the previous literature, which we take to be exemplified by MacArthur's $K$-selection argument \citep{macarthur_1967}.

MacArthur considered a population with two types that have densities $n_1$ and $n_2$ subject to density-dependent growth,
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2). \label{eq:macgeneral}
\end{equation}
The environment is assumed to remain constant apart from changing type densities. The functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are sufficiently large, because the resources required for growth are limited. This defines nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space. The outcome of selection is then determined by the relationship between these nullclines. Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline. Thus, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).

MacArthur used ``$K$'' to label the four intersection points of the nullclines with the axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(K_{21},0)=0$ and $f_2(0,K_{22})=0$. These $K$ values determine whether a region of higher-density growth exists for each type, provided that the nullclines are close to being straight lines. Note that only $K_{11}$ and $K_{22}$ are equilibrium densities akin to the $K$ parameter in the logistic model (Fig.~\ref{fig:Ksel}a). The other intersection points, $K_{12}$ and $K_{21}$, are related to competition between types. To be more concrete, in the Lotka-Volterra competition model we have
\begin{align}
f_1(n_1,n_2) = r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1\nonumber\\
f_2(n_1,n_2) = r_2(1-\alpha_{22}n_1-\alpha_{21}n_2)n_2\label{eq:LV}
\end{align}
where $\alpha_{11}=1/K_{11}$ and $\alpha_{22}=1/K_{22}$ measure competitive effects within types, while $\alpha_{12}=1/K_{12}$ and $\alpha_{21}=1/K_{21}$ measure competitive effects between types. Hence, ``fitness is $K$'' in crowded populations \citep[pp. 149]{macarthur_1967} in the sense that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards). This general idea is much broader than selection for greater equilibrium density \citep{gill_1974}.

\begin{figure}
\centering
\includegraphics[scale=0.8]{LVvslottery.pdf}
\caption{\label{fig:LVvslottery} Selection between types with identical equilibrium density but different inter-type competitive ability. (a) Lotka-Volterra competition (Eq.~\ref{eq:LV}) with $r_1=r_2=1$, $\alpha_{11}=\alpha_{22}=1$, $\alpha_{12}=0.9$ and $\alpha_{21}=1.2$. Trajectories do not follow the line $N=K_{11}=K_{22}$. (b) Lottery competition (Eq.~\ref{eq:master}) with $b_1=b_2=5$, $d_1=d_2=1.1$ and $c_1/c_2=5$. Trajectories converge on the line $N=K_{11}=K_{22}$.}
\end{figure}

Compared to simple birth-death models \citep{kostitzin_1939} or variants of the logistic \citep{roughgarden_1979}, the Lotka-Volterra model clearly distinguishes between intra- and inter-type competitive effects. Thus, when selection acts on inter-type competitive effects, one type can displace another without having a greater equilibrium density (Fig.~\ref{fig:LVvslottery}a). This has been termed ``$\alpha$-selection'' to distinguish it from $K$-selection, which involves intra-type competitive effects and changes in equilibrium density \cite{gill_1974,joshi_2001}. Although the initial and final densities of an $\alpha$-selection sweep are the same, density nevertheless does change transiently in the Lotka-Volterra model (constant density only occurs for a highly restricted subset of $r$ and $\alpha$ values; further details in Appendix C; also see \citealt{mallet_2012,smouse_1976}). Intuitively, for one type to exclude the other, competitive suppression of growth between types must be stronger than competitive suppression of growth within types, causing $N$ to dip over a sweep (Fig.~\ref{fig:LVvslottery}a).

In contrast to both $K$ and $\alpha$ selection, density trajectories for $c$-selection in our variable-density lottery converge on a line of constant equilibrium density (Fig.~\ref{fig:LVvslottery}b). This means that once $N$ reaches demographic equilibrium, selective sweeps behave indistinguishably from a constant-$N$ relative fitness model (Fig.~\ref{fig:Ksel}b). Thus, for $c$ sweeps, the selection factor $(W_2-W_1)/\overline{W}$ in Eq.~\eqref{eq:canonicaldisc} depends on frequency only, not density, provided that nothing else pushes $N$ out of demographic equilibrium over the sweep. This uncoupling of density from ongoing $c$-selection arises due to the presence of an excess of propagules which pay the cost of selection without affecting adult density \citep{nei1971fertility}.

@@

For the relative fitness model Eq.~\eqref{eq:canonicaldisc} to break down, the selection factor $(W_2-W_1)/\overline{W}$ must depend on density. As shown in Fig.~\ref{fig:DDS_lottery}, $(W_2-W_1)/\overline{W}$ is independent of $N$ in the case of $d$-selection. Selection is also independent of $N$ when the population is at demographic equilibrium and $N$ is unaffected by ongoing selection; as is the case for $c$-selection. Thus, to invalidate Eq.~\eqref{eq:canonicaldisc}, we require that selection is density-dependent, and also that density is changing. This can obviously occur if density-dependent selection happens in a population far from demographic equilibrium, in which case the validity of Eq.~\eqref{eq:canonicaldisc} depends on the specifics of the rate and magnitude of demographic change (we return to this in the Discussion). However, Eq.~\eqref{eq:canonicaldisc} can break down even in demographically-stable populations if a density-regulating trait is subject to density-dependent selection, as is the case for the $b$ trait in our variable-density lottery.

Before we discuss the $b$ trait, it is helpful to summarize how density-dependent selection on a density-regulating trait can invalidate Eq.~\eqref{eq:canonicalcont} in simpler continuous-time models. This applies, for example, to $K$-selection in the logistic \citep{kimura1969natural,crow_1970}. We consider the simple birth-death model \citep{kostitzin_1939} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i, \label{eq:simplebirthdeath}
\end{equation}
where $\delta_i$ is per-capita mortality due to crowding (for simplicity, there are no deaths when uncrowded). Starting from a type $1$ population in equilibrium, a variant with $\delta_2=\delta_1(1-\epsilon)$ has density-dependent selection coefficient $s=\epsilon \delta_1 N$ in Eq.~\eqref{eq:canonicalcont}, which will change over the course of the sweep as $N$ shifts from its initial type $1$ equilibrium to a type $2$ equilibrium. From Eq.~\eqref{eq:simplebirthdeath}, the equilibrium densities at the beginning and end of the sweep are $N_{\rm initial}=b_1/\delta_1$ and $N_{\rm final}=b_1/(\delta_1(1-\epsilon))=N_{\rm initial}/(1-\epsilon)$ respectively, and so $s_{\rm initial}= \epsilon b_1$ and $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$. Consequently, substantial deviations from Eq.~\eqref{eq:canonicalcont} occur if there is sufficiently strong selection on $\delta$ (Fig.~\ref{fig:strengthofselection}).

\begin{figure}
\centering
\includegraphics[scale=0.8]{strengthofselection.pdf}
\caption{\label{fig:strengthofselection} (a) Change in the selection coefficient between the beginning and end of a sweep of a type that experiences proportionally $1-\epsilon$ fold fewer crowding-induced deaths. The population is in demographic equilibrium at the start and end of the sweep. (b) Example equilibrium-to-equilibrium sweep.}
\end{figure}

Equilibrium-to-equilibrium $b$-sweeps in our variable-density lottery are qualitatively different from $\delta$ sweeps in this simpler birth-death model. Greater $b$ means not only that more propagules contest the available territories, but also that a greater fraction of unoccupied territories receive propagules. Together, the net density-dependent effect on $b$-selection is negligible: in a single-type equilibrium we have $W_i=1$ and $b_i/\overline{b}=1$, and hence the density-dependence factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ in Eq.~\eqref{eq:masterequalc} has the same value $d_i-1$ at the beginning and end of a $b$-sweep (recall that $\frac{W_2-W_1}{\overline{W}}=\frac{f(\overline{b},N)}{1+f(\overline{b},N)}\frac{b_2-b_1}{\overline{b}}$ for $b$-selection). During the sweep there is some deviation in $f(\overline{b},N)$, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the density-dependent deviation in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in our variable-density lottery is only of order $\epsilon^2$; see Appendix D for details). Since selection must already be strong for a $\delta$-sweep to invalidate Eq.~\eqref{eq:canonicalcont}, the density-independent model applies effectively exactly for equilibrium $b$-sweeps (Fig.~\ref{fig:bsweep}).

>>>>

Relative fitness is often parameterized in terms of selection coefficients which represent the advantages of different types relative to each other. For instance, in continuous time $s=r_2-r_1$ is the selection coefficient of type $2$ relative to type $1$. Assuming that only $2$ and $1$ are present, the change in frequency can be written as
\begin{equation}
\frac{d p_2}{dt}=s p_2 (1-p_2). \label{eq:canonicalcont}
\end{equation}
Thus, if $r_1$ and $r_2$ are constant, the frequency of the second type will grow logistically with a constant rate parameter $s$. We then say that selection is independent of frequency and density. The discrete time case is more complicated. Defining the selection coefficient by $W_2=(1+s)W_1$, and again assuming $1$ and $2$ are the only types present, we have  
\begin{equation}
\Delta p_2=\frac{W_2-W_1}{\overline{W}} p_2 (1-p_2)=\frac{s}{1+sp_2} p_2 (1-p_2). \label{eq:canonicaldisc}
\end{equation}
We will refer to both the continuous and discrete time selection equations \eqref{eq:canonicalcont} and \eqref{eq:canonicaldisc} throughout this paper, but the simpler continuous time case will be our point of comparison for the rest of this section.

In a constant environment, and in the absence of crowding, $r_i$ is a constant ``intrinsic'' population growth rate. The interpretation of Eq.~\eqref{eq:canonicalcont} is then simple: the selection coefficient $s$ is simply the difference in intrinsic growth rates. However, growth cannot continue at a non-zero constant rate indefinitely: the population is not viable if $r_i<0$, whereas  $r_i>0$ implies endlessly increasing population density. Thus, setting aside unviable populations, the increase in population density must be checked by crowding. This implies that the Malthusian parameters $r_i$ eventually decline to zero (e.g.~\citealt[pp. 203]{begon_1990}). Selection can then be density-dependent, and indeed this is probably not uncommon, because crowded and uncrowded conditions can favor very different traits \citep{travis_2013}. Eq.~\eqref{eq:canonicalcont} is then not a complete description of selection --- it lacks an additional coupled equation describing the dynamics of $N$, on which $s$ in Eq.~\eqref{eq:canonicalcont} now depends. In general we cannot simply specify the dynamics of $N$ independently, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{travis_2013}. Thus, in the presence of density-dependent selection, the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates. Note that frequency-dependent selection does not raise a similar problem, because a complete description of selection still only requires us to model the type frequencies, not the ecological variable $N$ as well.

In practice, many population genetic models simply ignore density dependence and assign a constant relative fitness to each type. Selection is typically interpreted as operating through viability, but the ecological processes underlying the regulation of population density are frequently left unspecified (e.g. \citealt{gillespie_2004}; \citealt{nagylaki_1992}; \citealt{ewens_2004}). Density either does not enter the model at all, or if finite-population size effects (``random genetic drift'') are important, then $N$ is typically assumed to have reached some fixed equilibrium value (Fig.~\ref{fig:Ksel}b; for some approaches to relaxing the constant $N$ assumption in finite populations, see \citealt{lambert_2005,parsons_2007,chotibut_2017,constable_2017}).

@@

The constant-$N$, constant-$s$ description of selection also limits consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction and trait evolution. A variety of approaches have been developed to address this in quantitative genetics \citep{burger1995evolution,engen_2013}, population genetics \citep{bertram2017predicting} and adaptive dynamics \citep{ferriere2013eco,dieckmann2004adaptive}. Although density-dependent selection is pertinent to these longer-term issues, our focus here is the description of the time-dependent process by which selection changes allele frequencies. This is particularly critical for making sense of evolution at the genetic level, for which we now have abundant data.

@@

Here we argue that the simplified models of density-regulated growth mentioned above are potentially misleading in their representation of the interplay between selection and density. This ultimately derives from their failure to account for ``reproductive excess'', that is, an excess of juveniles that experience stronger selection than their adult counterparts \citep{turner1968population}. By allowing selection to be concentrated at a juvenile ``bottleneck'', reproductive excess makes it possible for the density of adults to remain constant even under strong selection. Reproductive excess featured prominently in early debates about the regulation of population density (e.g. \citealt{nicholson_1954}), and also has a long history in evolutionary theory, particularly related to Haldane's ``cost of selection'' \citep{haldane_1957,turner1968population}. Additionally, reproductive excess is implicit in foundational evolutionary-genetic models like the Wright-Fisher, where each generation involves the production of an infinite number of zygotes, of which a constant number $N$ are sampled to form the next generation of adults. Likewise in the Moran model, a juvenile is always available to replace a dead adult every iteration no matter how rapidly adults are dying, and as a result $N$ remains constant.

@@

In models with detailed age structure, it is often assumed that the density of a ``critical age group'' mediates the population's response to crowding \citep[pp. 54]{charlesworth_1994}. Reproductive excess is a special case corresponding to a critical pre-reproductive age group. A central result of the theory of density-regulated age-structured populations is that selection proceeds in the direction of increasing equilibrium density in the critical age group \citep[pp. 148]{charlesworth_1994}. This is a form of the classical $K$-selection ideas discussed above, but restricted to the critical age group (juveniles, in this case). The interdependence of pre-reproductive selection and reproductive density is thus overlooked as a result of focusing on density in the critical age group.

@@

We restrict our attention to asexual haploids, since it is then clearer how the properties of selection are tied to the underlying population ecological assumptions. We assume that reproductively mature individuals (``adults'') require their own territory to survive and reproduce. All territories are identical, and the total number of territories is $T$. Time advances in discrete iterations, each representing the time from birth to reproductive maturity. In a given iteration, the number of adults of the $i$'th type will be denoted by $n_i$, the total number of adults by $N=\sum_i n_i$, and the number of unoccupied territories by $U=T-N$. We assume that the $n_i$ are large enough that stochastic fluctuations in the $n_i$ (drift) can be ignored, with $T$ also assumed large to allow for low type densities $n_i/T\ll 1$.

@@

We assume that adults cannot be ousted by juveniles, so that recruitment to adulthood occurs exclusively in unoccupied territories. When multiple propagules land on the same unoccupied territory, the winner is determined by lottery competition: type $i$ wins a territory with probability $c_i x_i/\sum_i c_i x_i$, where $c_i$ is a constant representing relative competitive ability (Fig. \ref{fig:lottery}). Since the expected fraction of unoccupied territories with propagule composition $x_1,\ldots,x_G$ is $p_1(x_1)\cdots p_G(x_G)$ where $G$ is the number of types present, and type $i$ is expected to win a proportion $c_i x_i/\sum_i c_i x_i$ of these, type $i$'s expected territorial acquisition is given by
\begin{equation}
\Delta_+ n_i=U\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_i c_i x_i} p_1(x_1)\cdots p_G(x_G). \label{eq:growthsumuncoupled}
\end{equation}
Here the sum only includes territories with at least one propagule present. Note that $\Delta_+ n_i$ denotes the \textit{expected} territorial acquisition. Fluctuations about $\Delta_+ n_i$ (i.e.~drift) will not be analyzed in this manuscript. Note that drift can become important if $U$ is not sufficiently large even though $n_i$ and $T$ are large (by assumption); we do not consider this scenario on biological grounds, since it implies negligible population turnover.

@@

Here we evaluate the expectation in Eq.~\eqref{eq:growthsumuncoupled} to better understand the dynamics of density-dependent lottery competition. Similarly to the classic lottery model, we replace the $x_i$, which take different values in different territories, with ``effective'' mean values. However, since we want to allow for low propagule densities, we cannot simply replace the $x_i$ with the means $l_i$ as in the classic lottery. For a low density type, growth comes almost entirely from territories with $x_i=1$, for which its mean density $l_i\ll 1$ is not representative. We therefore separate Eq.~\eqref{eq:growthsumuncoupled} into $x_i=1$ and $x_i>1$ components, taking care to ensure that the effective mean approximations for these components are consistent with each other (details in Appendix A). The resulting variable-density approximation only requires that there are no large discrepancies in competitive ability (i.e. we do not have $c_i/c_j\gg 1$ for any two types). We obtain
\begin{equation}
\Delta_+ n_i\approx \left[e^{-L}+(R_i+A_i)\frac{c_i}{\overline{c}}\right]l_i U, \label{eq:master}
\end{equation}
where
\begin{equation}
R_i=\frac{\overline{c}e^{-l_i}(1-e^{-(L-l_i)})}{c_i +\frac{\overline{c}L- c_il_i}{L-l_i}\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}},\nonumber \label{eq:Dr}
\end{equation}
and
\begin{equation}
A_i=\frac{\overline{c}(1-e^{-l_i})}{\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}c_il_i+\frac{\overline{c}L- c_il_i}{L-l_i}\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right)}. \nonumber \label{eq:Da}
\end{equation}

@@

We now turn to the issue of how density changes as a consequence of selection in our variable-density lottery, and in more familiar models of selection in density-regulated populations. In the latter, selection under crowded conditions typically induces changes in equilibrium density (see Introduction). In our variable-density lottery model, however, the competitive ability trait $c$ is not density-regulating, even though $c$ contributes to fitness under crowded conditions. Consequently, $c$-selection does not cause density to change. In this section we compare this $c$-selection behavior with the previous literature, which we take to be exemplified by MacArthur's $K$-selection argument \citep{macarthur_1967}.

MacArthur considered two types (with densities $n_1$ and $n_2$) in a constant environment subject to density-dependent growth,
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2). \label{eq:macgeneral}
\end{equation}
The outcome of selection is determined by the relationship between the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$. Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline.

MacArthur used the four intersection points of the nullclines with the axes, defined by $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(K_{21},0)=0$ and $f_2(0,K_{22})=0$, to analyze each type's exclusion or persistence. Note that only $K_{11}$ and $K_{22}$ are equilibrium densities akin to the $K$ parameter in the logistic model; the other intersection points, $K_{12}$ and $K_{21}$, are related to competition between types. For instance, in the Lotka-Volterra competition model we have
\begin{align}
f_1(n_1,n_2) = r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1\nonumber\\
f_2(n_1,n_2) = r_2(1-\alpha_{22}n_1-\alpha_{21}n_2)n_2\label{eq:LV}
\end{align}
where $\alpha_{11}=1/K_{11}$ and $\alpha_{22}=1/K_{22}$ measure competitive effects within types, while $\alpha_{12}=1/K_{12}$ and $\alpha_{21}=1/K_{21}$ measure competitive effects between types. Hence, ``fitness is $K$'' in crowded populations \citep[pp. 149]{macarthur_1967} in the sense that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards; \citealt{gill_1974}). However, even if the initial and final densities of an adaptive sweep in the Lotka-Volterra model are the same, $N$ nevertheless does change transiently (Fig.~\ref{fig:LVvslottery}a). Constant-$N$ over a sweep only occurs for a highly restricted subset of $r$ and $\alpha$ values (Appendix B; \citealt{mallet_2012,gill_1974,smouse_1976}).

\begin{figure}
\centering
\includegraphics[scale=0.8]{LVvslottery.pdf}
\caption{\label{fig:LVvslottery} Selection between types with identical equilibrium density but different inter-type competitive ability. (a) Lotka-Volterra competition (Eq.~\ref{eq:LV}) with $r_1=r_2=1$, $\alpha_{11}=\alpha_{22}=1$, $\alpha_{12}=0.9$ and $\alpha_{21}=1.2$. Trajectories do not follow the line $N=K_{11}=K_{22}$. (b) Lottery competition (Eq.~\ref{eq:master}) with $b_1=b_2=5$, $d_1=d_2=1.1$ and $c_1/c_2=5$. The lottery model nullclines are defined by $W_1=1$ (lower nullcline) and $W_2=1$ respectively in Eq.~\eqref{eq:absfit}. For the discrete-time Eq.~\eqref{eq:master}, trajectories are the flow lines of the vector field obtained by evaluating the direction of the local changes in $n_1$ and $n_2$; these converge on the line $N=K_{11}=K_{22}$.}
\end{figure}

In contrast, density trajectories for $c$-selection in our variable-density lottery converge on a line of constant equilibrium density (Fig.~\ref{fig:LVvslottery}b). This means that once $N$ reaches demographic equilibrium, selective sweeps behave indistinguishably from a constant-$N$ relative fitness model(Fig.~\ref{fig:Ksel}b). Thus, for $c$-sweeps in a constant environment, the selection factor $(W_2-W_1)/\overline{W}$ in Eq.~\eqref{eq:canonicaldisc} is density-independent. This uncoupling of density from $c$-selection arises due to the presence of an excess of propagules which pay the cost of selection without affecting adult density \citep{nei1971fertility}.

@@

For density to matter in Eq.~\eqref{eq:canonicaldisc}, selection must be density-dependent and density must be changing. This can occur in a constant environment if selection acts on a density-regulating trait. Consider the simple birth-death model \citep{kostitzin_1939} 
\begin{equation}
\frac{d n_i}{dt}=(b_i -\delta_iN) n_i, \label{eq:simplebirthdeath}
\end{equation}
where $\delta_i$ is per-capita mortality due to crowding. Starting from a type $1$ population in equilibrium, a variant with $\delta_2=\delta_1(1-\epsilon)$ has density-dependent selection coefficient $s=\epsilon \delta_1 N$, which will change over the course of the sweep as $N$ shifts from its initial type $1$ equilibrium to a type $2$ equilibrium. The equilibrium densities at the beginning and end of the sweep are $N_{\rm initial}=b_1/\delta_1$ and $N_{\rm final}=b_1/(\delta_1(1-\epsilon))=N_{\rm initial}/(1-\epsilon)$ respectively, and so $s_{\rm initial}= \epsilon b_1$ and $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$. Consequently, substantial deviations from Eq.~\eqref{eq:canonicalcont} occur if there is sufficiently strong selection on $\delta$ (Fig.~\ref{fig:strengthofselection}; \citealt{kimura_1978,kimura1969natural}).

\begin{figure}
\centering
\includegraphics[scale=0.8]{strengthofselection.pdf}
\caption{\label{fig:strengthofselection} (a) Change in the selection coefficient between the beginning and end of a sweep of a type that experiences proportionally $1-\epsilon$ fold fewer crowding-induced deaths. The population is in demographic equilibrium at the start and end of the sweep.}
\end{figure}

In our variable density lottery, $b$ regulates density and is subject to density-dependent selection, yet $b$-sweeps are qualitatively different from $\delta$ sweeps in the above example. Greater $b$ means not only that more propagules contest the available territories, but also that a greater fraction of unoccupied territories receive propagules. Together, the net density-dependent effect on $b$-selection is negligible: in a single-type equilibrium we have $W_i=1$ and $b_i/\overline{b}=1$, and hence the density-dependence factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ in Eq.~\eqref{eq:masterequalc} has the same value $d_i-1$ at the beginning and end of a $b$-sweep (recall that $\frac{W_2-W_1}{\overline{W}}=\frac{f(\overline{b},N)}{1+f(\overline{b},N)}\frac{b_2-b_1}{\overline{b}}$ for $b$-selection). During the sweep there is some deviation in $f(\overline{b},N)$, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the density-dependent deviation in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in our variable-density lottery is only of order $\epsilon^2$; see Appendix C for details). Since selection must already be strong for a $\delta$-sweep to invalidate Eq.~\eqref{eq:canonicalcont}, the density-independent model applies almost exactly for equilibrium $b$-sweeps (Fig.~\ref{fig:bsweep}).
